---
title: "\"Expected Returns\" Project Replication Examples"
subtitle: (draft)
author: "Vito Lestingi"
date: "`r Sys.Date()`"
abstract: "This document includes replication material on some academic and practitioners' literature instrumental for our RGSoC 2020 project. The document itself is meant to be completely reproducible."
output: 
  rmarkdown::pdf_document:
    citation_package: biblatex
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{\"Expected Returns\" Project Replication Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

```{r, echo=FALSE}
# Mentors, please feel free to add yourself to the authors' field if you wish. 
```

# Introduction
Our main reference is \textcite{ilmanen-2011} in that we want to reproduce selected key results the author discusses on factor-based investing. However, to achieve those results many more studies from the academic and practitioners' literature need to be implemented first. An excellent resource we are using to have a shared, quantitative modeling oriented, reference is \textcite{engle-2016}.

# Fundamental factor models

## Three-factor model
This model was introduced by \textcite{fama-french-1993}, expanding the work begun in \textcite{fama-french-1992}. In their cross-sectional analysis, the authors use the method of \textcite{black-jensen-scholes-1972}. However, the illustration \textcite{engle-16} do of the model applies the usual \textcite{fama-macbeth-1973} regression analysis. I suspect this difference is due to the fact that, whereas the former work is aimed at inferring and comparing common premia between both stocks and bonds, the latter is focused on providing an analysis on stock returns.

Following \textcite{engle-16}, we express the so called Fama-French *three-factor model* as
$$
    r_{p,t} = \beta_{0} + \beta_{1, p}MKT_{t} + \beta_{2, p}SMB_{t} + \beta_{3, p}HML_{t} + \epsilon_{t}
$$
where $r_{p,t}$ is the excess return of the portfolio $p$ over the time period $t$ (or, similarly, $r_{i,t}$ for a given security); $MKT_{t}$, $SMB_{t}$, and $HML_{t}$ are the returns of the "market", "small minus big" and "high minus low" factor mimicking portfolios, respectively. The intercept $\beta_{0}$ captures the average excess return that remain unexplained by the factors considered, whereas the remaining betas represent the sensitivities with respect to the each corresponding factor. The variable $\epsilon_{t}$ is the error term.

# Estimating factor models



```{r, echo=FALSE, results='hide'}
# Following suggestions, I'm working on reproducing models with existing functionality first.
# In particular, the 'FactorAnalytics' package has been suggest as useful. Which likely it is.
# However, I'm encountering some small issues in first attempts.
# 
# TODO:
# There are many aspects overlooked here. They must be included in our implementations 
# to guarantee analysis accuracy and usage flexibility (e.g. data winsorization).
# The examples below are first attempts to get the mechanics of some 'FactorAnalytics' 
# functions and evaluate their feasibility for our purposes.
# 

# install_github('FactorAnalytics')
library(FactorAnalytics)
library(xts)

### Experimenting fitFfm() to fit FF three-factor model ###

## Get and set data to pass to fitFfm()
data("factorDataSetDjia5Yrs") # don't run upon knitting
# select some variables needed
factorDataSetDjia5Yrs <- factorDataSetDjia5Yrs[, c('DATE', 'TICKER', 'NAME', 'RETURN.OLD', 'RETURN', 'RETURN.DIFF')]
ff3.data.monthly <- GetFactors('FF3', 'FF', freq='monthly') # don't run upon knitting

# make 'ff3.data.monthly' and 'factorDataSetDjia5Yrs' compatible for testing
ff3.data.monthly.dates <- index(ff3.data.monthly)
ff3.data.monthly <- as.data.frame(coredata(ff3.data.monthly))
ff3.data.monthly <- cbind(ff3.data.monthly.dates, ff3.data.monthly)
colnames(ff3.data.monthly)[1] <- 'DATE'
factorDataSetDjia5Yrs$DATE <- as.Date(factorDataSetDjia5Yrs$DATE)

test.data <- merge(factorDataSetDjia5Yrs, ff3.data.monthly)

# excess returns
test.data$EXC.RETURN <- test.data$RETURN - test.data$RF # FF risk-free rate (one-month Treasury bill rate from Ibbotson Associates)
test.data$RF <- NULL

# factor exposures
exposure.vars <- c('MKT.RF', 'SMB', 'HML')

## Run fitFfm() to fit factor model using cross-sectional regression
fit.test <- fitFfm(data=test.data, asset.var="TICKER", ret.var="EXC.RETURN", date.var="DATE", 
                   exposure.vars=exposure.vars, addIntercept=TRUE)
fit.test

```




```{r plain Fama-MacBeth, echo=FALSE, results='hide'}
# 
# TODO:
# Again, there are aspects overlooked. Both in the pre-processing (winsorization,
# truncation, etc.) and in the post-processing of the Fama-MacBeth methodology 
# (standard error corrections, test statistics adjustements and so on). Then their
# summaries as well.
# Once the core method is in place they are both straightforward.
# 
# Also, it's rather naive in this form. However, it can be more or less easily 
# improved and extended to other models. 
# 

library(FactorAnalytics) # for stock data

# Import data

# Get stocks data
data("factorDataSetDjia5Yrs")
factorDataSetDjia5Yrs <- factorDataSetDjia5Yrs[, c('DATE', 'TICKER', 'RETURN')]
factorDataSetDjia5Yrs$DATE <- as.Date(factorDataSetDjia5Yrs$DATE)

# Get FF3 factors data
ff3.data.monthly <- GetFactors('FF3', freq='monthly')
ff3.data.monthly.dates <- index(ff3.data.monthly)
ff3.data.monthly <- as.data.frame(coredata(ff3.data.monthly))
ff3.data.monthly <- cbind(ff3.data.monthly.dates, ff3.data.monthly)
colnames(ff3.data.monthly)[1] <- 'DATE'

# Merge & match FF3 factors data and stocks data
test.data <- merge(factorDataSetDjia5Yrs, ff3.data.monthly)

## Fama-MacBeth for FF Three-factor model, looping over lm()
test.data$EXC.RETURN <- test.data$RETURN - test.data$RF # FF risk-free rate (one-month Treasury bill rate from Ibbotson Associates)
periods <- unique(factorDataSetDjia5Yrs$DATE)
tickers <- unique(factorDataSetDjia5Yrs$TICKER)
factor.vars <- c('MKT.RF', 'SMB', 'HML')

# Run N time-series regressions
beta.coefs <- matrix(NA, length(tickers), length(factor.vars) + 1)
for (i in 1:length(tickers)) {
  asset.reg <- lm(EXC.RETURN ~ MKT.RF + SMB + HML, data=test.data[which(test.data[, 'TICKER'] == tickers[i]), ])
  beta.coefs[i, ] <- asset.reg$coefficients
}
row.names(beta.coefs) <- tickers
colnames(beta.coefs) <- c('(Intercept)', 'MKT.RF', 'SMB', 'HML')
beta.coefs

# Run T cross-sectional regressions 
gamma.coefs <- matrix(NA, length(periods), length(factor.vars) + 1)
gamma.rsq <- matrix(NA, length(periods), 1)
for (t in 1:length(periods)) {
  per.period.assets.ret <- test.data[which(test.data[, 'DATE'] == periods[t]), 'EXC.RETURN']
  data <- data.frame('EXC.RETURN'=per.period.assets.ret, beta.coefs[, 2:4]) # excluding prev. est. intercepts
  time.reg <- lm(EXC.RETURN ~ MKT.RF + SMB + HML, data=data)
  gamma.coefs[t, ] <- time.reg$coefficients
}
gamma.coefs <- data.frame(periods, gamma.coefs)
colnames(gamma.coefs) <- c('DATE', '(Intercept)', 'MKT.RF', 'SMB', 'HML')
gamma.coefs

# Average last regression coefficients estimates
avg.coefs <- colMeans(gamma.coefs[, 2:ncol(gamma.coefs)])
avg.coefs
```

