---
title: "\"Expected Returns\" Project Replication Examples"
subtitle: (draft)
author: "Vito Lestingi"
date: "`r Sys.Date()`"
abstract: "This document includes replication material on some academic and practitioners' literature instrumental for our RGSoC 2020 project. The document itself is meant to be completely reproducible."
output: 
  rmarkdown::pdf_document:
    citation_package: biblatex
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{\"Expected Returns\" Project Replication Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

```{r, echo=FALSE}
# Mentors, please feel free to add yourself to the authors' field if you wish. 
```

# Introduction
Our main reference is \textcite{ilmanen-2011} in that we want to reproduce selected key results the author discusses on factor-based investing. However, to achieve those results many more studies from the academic and practitioners' literature need to be implemented first. An excellent resource we are using to have a shared, quantitative modeling oriented, reference is \textcite{engle-2016}.

# Fundamental factor models

## Three-factor model
This model was introduced by \textcite{fama-french-1993}, expanding the work begun in \textcite{fama-french-1992}. In their cross-sectional analysis, the authors use the method of \textcite{black-jensen-scholes-1972}. However, the illustration \textcite{engle-16} do of the model applies the usual \textcite{fama-macbeth-1973} regression analysis. I suspect this difference is due to the fact that, whereas the former work is aimed at inferring and comparing common premia between both stocks and bonds, the latter is focused on providing an analysis on stock returns.

Following \textcite{engle-16}, we express the so called Fama-French *three-factor model* as
$$
    r_{p,t} =  \beta_{0} + \beta_{1, p}MKT_{t} + \beta_{2, p}SMB_{t} + \beta_{3, p}HML_{t} + \epsilon_{t}
$$
where $r_{p,t}$ is the excess return of the portfolio $p$ over the time period $t$ (or, similarly, $r_{i,t}$ for a given security); $MKT_{t}$, $SMB_{t}$, and $HML_{t}$ are the returns of the "market", "small minus big" and "high minus low" factor mimicking portfolios, respectively. The intercept $\beta_{0}$ captures the average excess return that remain unexplained by the factors considered, whereas the remaining betas represent the sensitivities with respect to the each corresponding factor. The variable $\epsilon_{t}$ is the error term.




