---
title: "Value-oriented equity selection"
subtitle: (draft)
author: "Vito Lestingi & Justin Shea"
date: "`r Sys.Date()`"
abstract: "This document includes replication material on some academic and practitioners' literature instrumental for our RGSoC 2020 project. The document itself is meant to be completely reproducible."
output: 
  rmarkdown::pdf_document:
    citation_package: biblatex
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{\"Expected Returns\" Project Replication Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

```{r, echo=FALSE}
# Mentors, please feel free to add yourself to the authors' field if you wish. 
```

# Introduction
 
Value intro

# Data

Get data from:
https://github.com/JustinMShea/ExpectedReturns/blob/master/inst/parsers/Devil-in-HML-Details.R

# Models

## Fama French HML Benchmark

Use the data above to run the standard FF model and save the results

## HML Devil in the Details

Now run a new model, and use the Devil Factors and save the results.

Compare the results

## Additional value factors

Inspired by the recent Cliff Assness discussion and AQR paper by the same title, 
Let's model the value factors he lists. We have some of these factors already
in FactorAnaltyics data sets.

https://www.aqr.com/Insights/Perspectives/Is-Systematic-Value-Investing-Dead


# Presenting results

Print resulting model estimates in a table, comparing estimates and fit. See Engle for inspiration.
