---
title: '"Value and Momentum Everywhere"'
subtitle: (draft)
author: "Vito Lestingi"
date: "`r Sys.Date()`"
abstract: "This document includes replication material on some academic and practitioners' literature instrumental for our RGSoC 2020 project. The document itself is meant to be completely reproducible."
output:
  rmarkdown::pdf_document:
    citation_package: biblatex
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{"Value and Momentum Everywhere"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)

library(xts)
library(plm)

```

```{r, echo=FALSE}
# Mentors, please feel free to add yourself to the authors' field if you wish.
```

# Introduction
In these notes we replicate \textcite{asness-moskowitz-pedersen-2013} (AMP hereafter).
The authors aim at providing a *global*, across both markets/asset classes and across countries, asset pricing approach. This is what "everywhere" in "Value and Momentum *Everywhere*" stands for.

## Data and Portfolio Construction
Factors and portfolios are constructed from January 1972 through July 2011. All data series have a monthly frequency. They construct zero-cost long/short VME factors and long-only value and momentum portfolios, sorted into tertiles, for both value and momentum in eight markets/asset classes.

Because of the *global* nature of their pricing approach, authors use as their $MKT$ factor an international portfolio proxy, represented by the MSCI *World Index*. The index has 1,637 large and mid cap constituents across 23 Developed Markets countries and, according to MSCI Inc., it "covers approximately 85% of the free float-adjusted market capitalization in each of them." We provide a comparison with the international global equity portfolio constructed by \textcite{asness-frazzini-2013}.

```{r AQR Global Aggregate Equity Portfolios vs MSCI Barra Indexes}
# Source AQR's Global Aggregate Equity Portfolio ("AQR GAP")
# NOTE: returns in decimal unit
source("Value--Devil-in-HMLs-Details.R")

# Subset AQR GAP data points 
from <- first(which(!is.na(HML_Devil.MKT$Global)))
to <- length(HML_Devil.MKT$Global)
MKT.GLOBAL <- xts(HML_Devil.MKT$Global[from:to], HML_Devil.MKT$DATE[from:to])
colnames(MKT.GLOBAL) <- 'MKT.GLOBAL'

# Correlation methods to pass to cor()
cor.methods <- c('pearson', 'spearman')
```

```{r AQR GAP vs MSCI WI}
# NOTE: Get 'MSCI.WI' from ExpectedReturns/inst/parsers/MSCI-WI.R
# NOTE: returns in decimal unit

# Merge and compute correlation(s)
global.proxies <- merge(MKT.GLOBAL, MSCI.WI$COMP.RET)
min.tp <- first(which(!is.na(global.proxies$MKT.GLOBAL))) + 1
max.tp <- nrow(global.proxies)
gap.wi.cor <- sapply(1:length(cor.methods), function(x) {
  as.numeric(
    cor(
      global.proxies$MKT.GLOBAL[min.tp:max.tp],
      global.proxies$COMP.RET[min.tp:max.tp],
      use='na.or.complete',
      method=cor.methods[x]
    )
  )
})
gap.wi.cor

# Plot comparison
# global.proxies.cont <- na.locf(global.proxies)
global.proxies.cont <- na.fill(global.proxies, c(NA, 'extend', 'extend'))
plot.xts(global.proxies.cont, main='AQR\'s Global MKT vs MSCI World Index',
         screens=factor(1, 1), col=c('black', 'blue'), lwd=c(0.95, 0.95))
addLegend(
  'bottomleft', 
  legend.names = c(
    'Global MKT', 'MSCI WI', 
    expression(paste(rho[Pearson], ' = 0.9834098')),
    expression(paste(rho[Spearman], ' = 0.9790861'))
  ), 
  cex=0.75, lty=1, lwd=1, col=c('black', 'blue', NA, NA)
)
```

Because the , containes more securities then the MSCI WI, here we also provide a comparison with the MSCI *All Countries World Index* (MSCI ACWI). The latter synthesizes 3,040 large and mid cap constituents across 23 Developed Markets
and 26 Emerging Markets countries and, according to MSCI Inc., it "covers approximately 85% of the global investable equity opportunity set."

```{r AQR GAP vs vs MSCI ACWI}
# NOTE: Get 'MSCI.ACWI' from ExpectedReturns/inst/parsers/MSCI-ACWI.R
# NOTE: returns in decimal unit

# Merge and compute correlation(s)
global.proxies <- merge(MKT.GLOBAL, MSCI.ACWI$COMP.RET)
min.tp <- first(which(!is.na(global.proxies$COMP.RET)))
max.tp <- nrow(global.proxies)
gap.acwi.cor <- sapply(1:length(cor.methods), function(x) {
  as.numeric(
    cor(
      global.proxies$MKT.GLOBAL[min.tp:max.tp],
      global.proxies$COMP.RET[min.tp:max.tp],
      use='na.or.complete',
      method=cor.methods[x]
    )
  )
})
gap.acwi.cor

# Plot comparison
# global.proxies.cont <- na.locf(global.proxies)
global.proxies.cont <- na.fill(global.proxies, c(NA, 'extend', 'extend'))
plot.xts(global.proxies.cont, main='AQR\'s Global MKT vs MSCI ACWI',
         screens=factor(1, 1), col=c('black', 'blue'), lwd=c(0.95, 0.95))
addLegend(
  'bottomleft', 
  legend.names = c(
    'Global MKT', 'MSCI ACWI', 
    expression(paste(rho[Pearson], ' = 0.9887671')),
    expression(paste(rho[Spearman], ' = 0.9840154'))
  ), 
  cex=0.75, lty=1, lwd=1, col=c('black', 'blue', NA, NA)
)
```

As shown by the high correlations among these global equity portfolios, in principle each one of them may serve as the market portfolio return proxy.


### Asset classes and countries
<!--
If I'll have time after coding, I'd make a detailed table here.
- Vito
-->
The markets/asset classes contemplated include four individual equity markets (stock selection) and four broad asset classes (asset allocation). The four stock selection markets are: U.S. equities (US), U.K. equities (UK), Continental Europe equities (EU), Japanese equities (JP). The four asset allocation classes are: global equity indices (EQ), currencies (FX), fixed income (FI), and commodity futures (CM).

They are global individual stocks, global equity indices, currencies, global government bonds, commodity futures (27 types by different underlying),  

While the countries are...

### Value and momentum measures
* **Value measure(s).** Differ by asset class as not every of them has a book value measure as stocks do, thus the book-to-market ratio cannot be derived in most classes.
* **Momentum measure.** It is same for all asset classes, i.e. the common measure of the past 12-month cumulative raw return on the asset skipping the most recent month's return.

## Relation to Macroeconomic and Liquidity Risk
This section of \textcite{asness-frazzini-pedersen-2013} blends very well with \textcite{ilmanen-2011}'s "Broader Themes", especially with what are commonly referred to as "macroeconomic factors" or simply "macros" in the financial industry jargon. Indeed, at a first sight it appears a more quantitative-oriented look at many ideas the former author already discussed.
In particular, the authors study the underlying economic sources driving value and momentum returns: 
* Long run consumption growth
* MSCI world index ("market")
* Recession dummy
* GDP growth
* Liquidity

The *Liquidity Risk Exposure* is particularly interesting to study and deserves particular attention. In this context, it is so especially from the point of view of market-level traded liquidity, given the ubiquitous influence of the factor for the entire economic system. 
It really has been in Economists' thoughts since far longer than the modern standard literature reminds us. Recently
it has been linked with financial crises (see what happened in 1987), flash crashes (2010), and vice versa is very often in policy makers' decisions. See, for example, monetary policies actions Central Banks adopt worldwide in this regard.
\textcite{asness-moskowitz-pedersen-2013} pay particular attention to global events and link them with liquidity shocks. Then, they investigate the relation between value and momentum returns and liquidity risk.

## Comovement and Asset Pricing Tests
```{r Get factors and portfolios data}
## VME data
# Until we establish a way to source in data, get data from here:
# - https://github.com/JustinMShea/ExpectedReturns/blob/master/inst/parsers/VME-Factors.R
# - https://github.com/JustinMShea/ExpectedReturns/blob/master/inst/parsers/VME-Portfolios.R

# Value and Momentum Everywhere Factors
VME.FACTORS <- VME.Factors[, c('DATE', 'VAL.EVR', 'MOM.EVR')]

# Get AMP Portfolios Returns and make them a panel
VME.PORTF.DATE <- VME.Portfolios$DATE
VME.Portfolios$DATE <- NULL
VME.PORTF.RET <- VME.Portfolios
portf.names <- colnames(VME.PORTF.RET)
t <- nrow(VME.PORTF.RET)
p <- ncol(VME.PORTF.RET)
portf.names <- as.data.frame(matrix(portf.names, nrow=t, ncol=p, byrow=TRUE))
VME.PORTF.RET <- asplit(VME.PORTF.RET, 2)
portf.names <- asplit(portf.names, 2)
VME.PORTF.RET <- Map(cbind, portf.names, VME.PORTF.RET)
VME.PORTF.RET <- Reduce(rbind, VME.PORTF.RET)
VME.PORTF.RET <- data.frame(PORTF=VME.PORTF.RET[, 1], RET=VME.PORTF.RET[, 2])
VME.PORTF.RET <- cbind(DATE=rep(VME.PORTF.DATE, p), VME.PORTF.RET)
VME.PORTF.RET$DATE <- as.Date(VME.PORTF.RET$DATE)
VME.PORTF.RET$RET <- as.numeric(VME.PORTF.RET$RET)

## Get Fama-French-Carhart Four Factors (MKT.RF, SMB, HML, MOM)
FF3 <- ExpectedReturns::GetFactors('FF3', freq='monthly')
MOM <- ExpectedReturns::GetFactors('MOM', freq='monthly')
min.tp <- max(first(index(FF3)), first(index(MOM)))
max.tp <- min(last(index(FF3)), last(index(MOM)))
tp <- paste(min.tp, max.tp, sep='/')
FF4 <- data.frame(
  DATE=seq.Date(min.tp, max.tp, by='month') - 1, # re-check B-o-M
  coredata(FF3[tp, ]),
  coredata(MOM[tp, ])
)

## Merge FF4, VME Factors and VME Portfolios Returns
data <- merge(VME.PORTF.RET, FF4, by='DATE')
data <- merge(data, VME.FACTORS, by='DATE')

## Calculate excess returns and merge
data$EXC.RET <- data$RET - data$RF
```

### Explaining Value/Momentum in One Market with Value/Momentum in Other Markets
$$
R_{i,t}^{p} - r_{f,t} = \alpha_{i}^{p} + \beta_{i}^{p}MKT_{t} + v_{i}^{p}\sum_{j \neq i}w_{j}VAL_{t}^{\textrm{everywhere}} + m_{i}^{p}\sum_{j \neq i}w_{j}MOM_{t}^{\textrm{everywhere}} + \epsilon_{i,t}^{p}
$$
where $w_{j}$ is the equal volatility weight for each asset class.

### Global Three-Factor Model

$$
R_{i,t}^{p} - r_{f,t} = \alpha_{i}^{p} + \beta_{i}^{p}MKT_{t} + v_{i}^{p}VAL_{t}^{\textrm{everywhere}} + m_{i}^{p}MOM_{t}^{\textrm{everywhere}} + \epsilon_{i,t}^{p}
$$
$R_{i,t}^{p}$ is the time $t$ return to portfolio $p$ among the six high, middle, and low value and momentum portfolios in one of the eight asset markets $i$, for a total of 48 test assets. 
Whereas, $VAL_{t}^{\textrm{everywhere}}$ and $MOM_{t}^{\textrm{everywhere}}$ are equal-volatility-weighted across asset-class value and momentum factors.

```{r Global Three-Factor Model}
## Prepare indexes to determine panel structure
# 'DATE.ID', each date gets same number id across
dates <- unique(data$DATE)
dates.id <- 1:length(dates)
dates.id <- data.frame('DATE'=dates, 'DATE.ID'=dates.id)
data <- merge(data, dates.id, by='DATE')
# 'PORTF.ID', each asset gets same number id across
portf <- unique(data$PORTF)
portf.id <- 1:length(portf)
portf.id <- data.frame('PORTF'=portf, 'PORTF.ID'=portf.id)
data <- merge(data, portf.id)
data <- data[order(data[, 'PORTF.ID'], data[, 'DATE.ID']), ]
row.names(data) <- NULL

# Time-series regressions
ts.reg <- plm::pmg(
  EXC.RET ~ MKT.RF + VAL.EVR + MOM.EVR, # TODO: only model for which we have data
  data=data, index=c('PORTF.ID', 'DATE.ID')
)
summary(ts.reg) # average coefficients
betas <- t(ts.reg$indcoef) # all coefficients
rownames(betas) <- portf
colnames(betas) <- paste('BETA', colnames(betas), sep='.')

# Cross-sectional regression
cs.data <- data[order(data[, 'DATE']), ]
cs.data <- data.frame(
  data[, c('DATE', 'EXC.RET', 'DATE.ID', 'PORTF.ID')], 
  betas[, 2:ncol(betas)]
)
cs.reg <- plm::pmg(
  EXC.RET ~ BETA.MKT.RF + BETA.VAL.EVR + BETA.MOM.EVR, 
  data=cs.data, index=c('DATE.ID', 'PORTF.ID')
)
summary(cs.reg) # average coefficients
gammas <- cs.reg$indcoef # all coefficients
gammas <- t(gammas)
gammas <- data.frame(dates, gammas, row.names=NULL)
colnames(gammas) <- c('DATE', '(Intercept)', colnames(betas)[-1])
```

```{r VME Table 6}
### TABLE 6, Panel A,  U.S. Asset Pricing Factors ###
# NOTE: Here the FF US MKT.RF factor is properly used
factors.vars <- c('MKT.RF', 'SMB', 'HML', 'MOM')
fv.seqs <- 1:length(factors.vars)
# Time-series regression by model
ts.reg.us <- lapply(fv.seqs, function(x) {
  # Models specifications
  model.formula <- formula(
    paste0(
      'EXC.RET ~ ',
      paste(factors.vars[1:x], collapse='+')
    )
  )
  plm::pmg(
    model.formula, data=data, 
    index=c('PORTF.ID', 'DATE.ID')
  )
})
# Models results
lapply(ts.reg.us, summary)
```

```{r VME Figure 6 plots}
### AMP Model Plot ###
VME.Portfolios <- apply(VME.Portfolios, 2, as.numeric)
amp.factors <- data[1:length(dates), c('MKT.RF', 'VAL.EVR', 'MOM.EVR')]
amp.factors.avg <- colMeans(amp.factors, na.rm=TRUE)
# Actual Average Returns 
# TODO: Full series lengths differ. Should use only obs in regressions
y <- colMeans(VME.Portfolios, na.rm=TRUE)
y <- y[sort(names(y))]
# Expected Returns
x <- betas %*% c(1, amp.factors.avg)
x <- x[sort(row.names(x)), ]
# Plot
plot(x=x, y=y, main='AMP Global Three-factor Model', 
     xlab='Portfolio Expected Returns', ylab='Portfolio Realized Average Returns', 
     pch=20, cex.lab=0.85)
lines(par()$usr[1:2], par()$usr[3:4], col='gray75')
text(x=x, y=y, names(x), cex=0.5, pos=4)
grid()
```

