---
title: "Value-oriented equity selection"
subtitle: (draft)
author: "Vito Lestingi & Justin Shea"
date: "`r Sys.Date()`"
abstract: "This document includes replication material on some academic and practitioners' literature instrumental for our RGSoC 2020 project. The document itself is meant to be completely reproducible."
output: 
  rmarkdown::pdf_document:
    citation_package: biblatex
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Value-oriented equity selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

```{r, echo=FALSE}
# Mentors, please feel free to add yourself to the authors' field if you wish. 
```

# Introduction
 
Value intro

# Data

Get data from:
https://github.com/JustinMShea/ExpectedReturns/blob/master/inst/parsers/Devil-in-HML-Details.R

# Models

## Fama French HML Benchmark

Use the data above to run the standard FF model and save the results

<!--
Have already re-elaborated and shared more or less all about Fama-French Three-factor 
variables definitions and portfolio construction from their website. We can just retrieve those here. 
This is one reason why that work was done and time invested in doing so extensively.
Time that now pays off.
- Vito
-->


## HML Devil in the Details
<!-- 
Now run a new model, and use the Devil Factors and save the results.
Compare the results. 
- Justin
-->

In the following we briefly review \textcite{asness-frazzini-2013} (AF hereafter) 
and reproduce this work thanks to replication data sets publicly available. Data 
are updated and maintained by AQR (www.aqr.com).
As will be clear from further discussions, for their analysis authors use data until
2010 or 2011, depending on the specific value measures being constructed, compared,
and analyzed.
Updated data sets go beyond the time frame originally covered in the paper, the 
most recent data are until May, 31, 2020. First observations dates vary by country,
the United States have the longer data series in the sample with the first observation 
reported on July, 31, 1926.

### Value measures
Introducing the value measures contemplated in this work, the first value measure 
studied is the standard practice pioneered by \textcite{fama-french-1992}. It is 
the most widely used measure in the literature and as such it is also referred to 
as the "standard method". 
The Fama-French value measure is defined
$$
bp_{t}^{a,l} = \ln(B/P_{fye})
$$
where $B$ is the book value per share (or market equity value) and $P_{fye}$ indicates
the price variable at the end of the fiscal year (rather than in June), both quantities
are expressed in local currency. Note that superscripts $a$ and $l$ indicate respectively
that this measure is annual (it is updated once a year) and lagged (prices are six 
to 18 months prior to date, not current prices). The $B/P$ ratio, at times denoted 
$BM$ is commonly known as the book-to-market ratio.

The second and third values measures, original of this work, are slight modifications 
of the standard method. As we will outline in further sections, these measures indeed 
resemble most of the authors research focus: should analysts or investors use lagged 
price in constructing valuation ratios?
The measures, expressed in local currency, are
$$
bp_{t}^{a,c} = \ln(B^*/P_t)
$$
and
$$
bp_{t}^{m,c} = \ln(B^*/P_t)
$$
The superscripts have the following meaning, the former is updated annually and 
computed using current prices ($c$ superscript), whereas the latter is update monthly 
($m$ superscript) and computed using most recent prices as per the June 30 rebalance 
date ($c$, for current, superscript).
In both the two above measures $B^*$ stands for the firms' book value per share 
(or market equity value) that has been adjusted for splits, dividends, and other 
corporate actions between fiscal year-end and portfolio formation dates. In other 
words, we can express this financial variable by means of an adjustment ratio:
$$
B^* = B\frac{Adj_t}{Adj_{fye}}
$$
with $Adj$ the cumulative adjustment factor, which adjusts between the fiscal 
year-end and the current date $t$.

All the three measures above are related to one another by the following relations,
$$
bp_{t}^{a,c} = bp_{t}^{a,l} - r_{fye,t}
$$
$$
bp_{t}^{m,c} = bp_{t}^{a,c} - r_{t,t+k}
$$
where 
$$
r_{t,s} = \ln(1 + R_{t,s})
$$
is the total log return between date $t$ and $t < s$. 

### Portfolio construction
<!--
I have already re-elaborated and shared more or less all about Fama-French Three-factor 
definitions and construction from their website. 
We can:
- simply retrieve those definitions here as well;
- discuss them only in the usual Fama-French model in its own section above (preferred).
Anyways, this is precisely why that method was used and time invested in doing so.
NOTE: This holds for U.S. calculation based factors, *not* for the International sample.
For the latter I'll add details and will be straightforward as well thereafter.
(see here meanwhile: http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html#International)

- Vito
-->
All portfolio returns are in USD, excess returns are returns above the U.S. Treasury bill rate.

The $HML^{devil}$ factor introduced by the authors is constructed from value-weighted portfolios with size/book-to-market breakpoints updated every calendar month. Portfolios are rebalanced every calendar month to maintain weights. The factor is defined as the average return on the two values portfolios minus the average return on the two growth portfolios, 
$$
HML^{devil} = \frac{1}{2}[(\textrm{Small Value} + \textrm{Big Value}) - (\textrm{Small Growth} + \textrm{Big Growth})]
$$


### What Proxies Best for the True Unobservable Book-to-market ratio?

The authors propose a less lagged version of the standard $BP_{t-1}^{a,l}$, which 
they define
$$
BP_{t-1}^{a,c} = \frac{B_{t-1}}{P_{t,(June)}}
$$

> "We show that using a more-current price is superior to the standard method of using prices at fiscal year-end as a proxy for the true B/P ratio, and superior in five-factor model regressions."
- Asness-Frazzini (2013, p. 49)

> "Does the fallen price make this more likely, less likely, or have no effect on whether this should be considered a value stock?
The answer depends on how much variation in B/P ratios is due to expected returns and how much is due to changes in future book values. Our findings show that true value stocks often show such price drops, and a measure that takes this fall into consideration, as our proposed method does, is superior to one that ignores it, as the standard method does."
- Asness-Frazzini (2013, pp. 49-50)

<!-- (DATA: where is book equity?)
To reproduce these results I need more data than currently available. 
Authors provide ME in the data sets you parsed, but BE is needed as well to get BMs.

Now, it seems to me that FactorsAnalytics' 'factorDataSetDjia5Yrs' data contains
P2B for DJIA companies so we may in principle recover data from there. 
But I don't think it's enough, it's not the same assets universe.
From the authors ("DATA, METHODOLOGY, AND TERMINOLOGY", p. 50):
"Our U.S. equity data includes all available common stocks on the merged CRSP/XpressFeed 
data between July 1950 and March 2011."

Doing some research I found that the great Prof. French shares BE data for the US.
(http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html#Breakpoints).
Couldn't find anything close to the "International sample".

- Vito
-->

<!-- (DATA: where is the short-term reversal factor?)
I can't find it from your parsed data set. Is it in the original data set?
Thankfully it can be conveniently be obtained with GetFactors() from Prof. French...
- Vito
-->

### Which Book-to-market ratio Form Better Value Portfolios? 

```{r Get AF (2013) Factors Data}
## Load libraries
library(xts)
library(plm)
library(FactorAnalytics)

## Run Justin's parser for Asness-Frazzini (2013) replication data set
https://github.com/JustinMShea/ExpectedReturns/blob/master/inst/parsers/Devil-in-HML-Details.R

## Get "The Devil in HML's Details" factors data from Justin's parser
# NOTE: "International Sample" assets data is missing, consider US only for now 
#       as we have some assets data available (see below)
# NOTE: In the paper sample period is 1950-2011 for US, 1983-2011 for Intl.
FACTORS.DATE <- HML_Devil.HML.DEV$DATE
MKT.RF <- HML_Devil.MKT$USA # TODO: is it net of RF already?
SMB <- HML_Devil.SMB$USA
HML <- HML_Devil.HML_FF$USA
HML.DEV <- HML_Devil.HML.DEV$USA
UMD <- HML_Devil.UMD$USA
REV.ST <- ExpectedReturns::GetFactors('REV', freq='monthly', term='ST') # they omit to provide it (see note 18)
STR <- data.frame(DATE=index(REV.ST) - 1, STR=coredata(REV.ST))         # recheck "arbitrary" B-o-M 
STR <- STR[which(STR$DATE %in% FACTORS.DATE), 'REV.ST']

TP <- max(length(FACTORS.DATE), length(REV.ST))

# TODO: STR decimal* or percentage?  
factors.data <- data.frame(
  DATE=FACTORS.DATE[1:TP],
  MKT.RF=MKT.RF[1:TP],
  SMB=SMB[1:TP],
  HML=HML[1:TP],
  UMD=UMD[1:TP],
  STR=STR[1:TP],
  HML.DEV=HML.DEV[1:TP]
)
```

```{r Get assets data}
# NOTE: We only have stocks data for the US available. Also, is monthly data.
#       So we can only attempt to replicate parts of the paper within these settings for now.
# NOTE: Authors have both a large assets universe and long time-series.
#       I picked "stocksCRSPscoresSPGMIraw" because, although it has less stocks 
#       than "Stock.df" (300 assets), it spans a longer period of time (1993-01-31 to 2015-12-31)
#       within that of interest.
data("stocksCRSPscoresSPGMIraw")
vars.keep <- c('Date', 'Ticker', 'Size', 'MPrc', 'MRetx') # assuming "MRetx" is stock excess returns
assets.data <- as.data.frame(stocksCRSPscoresSPGMIraw[, ..vars.keep])
colnames(assets.data) <- toupper(vars.keep)
```

```{r Merge Factors and Assets data series available}
# US sample observation dates in AF paper
US.SAMPLE.DATE.PAPER <- seq.Date(as.Date('1950-02-01'), as.Date('2012-01-01'), by='month') - 1
# Dates respecting all our constraints
dates <- as.Date(Reduce(intersect, list(factors.data$DATE, assets.data$DATE, US.SAMPLE.DATE.PAPER)))
# Merge
data <- merge(assets.data[which(assets.data$DATE %in% dates), ], factors.data)
```

```{r Replicate Part of Exhibit 6 (p. 57) with FactorAnalytics}
# TODO: which HML^{devil} is in the available data set, monthly & current? 
#       I think it is, they call it "our very timely HML^{m,c}". 
#       The others are the standard HML of FF (annual & lagged/current)

#############################################################
### Time-series regression with FactorAnalytics::fitFfm() ###
#############################################################
# exposure.vars <- c('MKT.RF', 'SMB', 'HML', 'UMD', 'STR', 'HML.DEV')
# fit.test.fa <- fitFfm(data=data, asset.var="TICKER", ret.var="MRETX", date.var="DATE", 
#                       exposure.vars=exposure.vars, addIntercept=TRUE)
# TODO: all 'MRETX' missing are for VLO during 1993--1997. Was it delisted?
# data[which(is.na(data[, 'MRETX'])), ]
```

```{r Replicate Part of Exhibit 6 (p. 57) with plm}
##############################################
### Time-series regression with plm::pmg() ###
##############################################
#
# Prepare indexes to determine panel structure
# 'PERIOD.ID', each date gets same number id across
periods <- unique(data$DATE)
periods.id <- 1:length(periods)
periods.id <- data.frame('DATE'=periods, 'PERIOD.ID'=periods.id)
data <- merge(data, periods.id, by='DATE')
# 'ASSET.ID', each asset gets same number id across
tickers <- unique(data$TICKER)
assets.id <- 1:length(tickers)
assets.id <- data.frame('TICKER'=tickers, 'ASSET.ID'=assets.id)
data <- merge(data, assets.id)
data <- data[order(data[, 'ASSET.ID'], data[, 'PERIOD.ID']), ] # sort data set
row.names(data) <- NULL
# Time-series regressions
fit.test.plm <- plm::pmg(MRETX ~ MKT.RF + SMB + HML + UMD + STR + HML.DEV, 
                         data=data, index=c('ASSET.ID', 'PERIOD.ID'))
summary(fit.test.plm)            # average coefficients
betas <- t(fit.test.plm$indcoef) # all coefficients
rownames(betas) <- tickers
colnames(betas) <- paste('BETA', colnames(betas), sep='.')
betas
```


### Value and momentum interaction
<!--
This is the reason last Saturday we were discussing to start from "Value and Momentum Everywhere". 
The same or very similar arguments, plus some of Ilmanen's "broader themes" (such as the liquidity factor), are discussed and analyzed in more depth. This fact is indicated by authors themselves, see p. 936 of "Value and Momentum Everywhere".
- Vito
-->
The paper also highlights important aspects on the dynamics between value and 
momentum strategies. In section 3 authors add that:
> "...in the presence of momentum, and for logical reasons having to do with the overlap of our value measure and the period used to form momentum, our more timely value measure also outperforms the more standard lagged measure."
- Asness-Frazzini (2013, p. 56)

## Additional value factors

Inspired by the recent Cliff Assness discussion and AQR paper by the same title, 
Let's model the value factors he lists. We have some of these factors already
in FactorAnaltyics data sets.

https://www.aqr.com/Insights/Perspectives/Is-Systematic-Value-Investing-Dead


# Presenting results

Print resulting model estimates in a table, comparing estimates and fit. See Engle for inspiration.
