---
title: "Value-oriented equity selection"
subtitle: (draft)
author: "Vito Lestingi & Justin Shea"
date: "`r Sys.Date()`"
abstract: "This document includes replication material on some academic and practitioners' literature instrumental for our RGSoC 2020 project. The document itself is meant to be completely reproducible."
output: 
  rmarkdown::pdf_document:
    citation_package: biblatex
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{\"Expected Returns\" Project Replication Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

```{r, echo=FALSE}
# Mentors, please feel free to add yourself to the authors' field if you wish. 
```

# Introduction
 
Value intro

# Data

Get data from:
https://github.com/JustinMShea/ExpectedReturns/blob/master/inst/parsers/Devil-in-HML-Details.R

# Models

## Fama French HML Benchmark

Use the data above to run the standard FF model with HML save the results

## HML Devil in the Details

Now run a new model, and use the Devil Factors 

Compare the results

## Additional value factors

Inspried by this, we will have to get this data and will cross that bridge when we get there.

https://www.aqr.com/Insights/Perspectives/Is-Systematic-Value-Investing-Dead


# Presenting results

Print resulting model estiamtes in a table, comparing estimates and fit. See Engle for inspiration.
