---
title: '"Which Factors? q-factor and traditional models"'
subtitle: "(draft)"
author: "Vito Lestingi"
date: "`r Sys.Date()`"
abstract: "This document includes replication material on some academic and practitioners' literature instrumental for our RGSoC 2020 project. The document itself is meant to be completely reproducible."
output:
  rmarkdown::pdf_document:
  citation_package: biblatex
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{"Which Factors? q-factor and traditional models"}
  %\VignetteEngine{rmarkdown::render}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_knit$set(
  root.dir = normalizePath('..')
)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  error = FALSE
)
```

# Introduction
These notes aim at reproducing several works on factors modeling and their comparisons. The framework is the one adopted by \textcite{hou-al-2018}, who systematically examine how seemingly different factor models are related among each other. In particular, their approach is focused on comparing their contributed q-factor models against an assorted set of emblematic models from various strains of the asset pricing literature, on both conceptual and empirical grounds.
These benchmark (fundamental) factor models include "standard" ones as the Fama-French five-factor model \parencite{fama-french-2015} and the Fama-French six-factor model \parencite{fama-french-2018}, models methodologically divergent from the Fama-French approach such as \textcite{stambaugh-yuan-2017} and \textcite{daniel-hirshleifer-sun-2020}, and also the "hybrid" model of \textcite{barillas-shanken-2018}. At the time of writing, we are focusing on analyses involving q-factor models and "traditional" models, that is the Fama-French models. This given latter authors' pioneering work in the field and their models widespread use as the de facto benchmarks in the broader literature.
In what follows we first of all introduce target models and their theoretical grounds, then proceed with spanning regressions for their statistical comparison and performance assessments.

# Data and methodology
Authors make available data for factors returns analyzed in their research at [global-q.org](http://global-q.org/index.html). Importantly, data sets are open-sourced and updated regularly. Similarly to the well known Fama-French academic library, data series they provide span several frequencies including daily, weekly (calendar, Friday close to Friday close), weekly (Wednesday-to-Wednesday, Wednesday close to Wednesday close), monthly, quarterly, and annual.

```{r Load data sets}
data(
  list=c(
    'FF5.monthly', 'FF6.monthly', 
    'Q4.monthly', 'Q5.monthly', 
    'SY4.monthly',
    'DHS3.monthly'
  )
)

# No out-of-the-box replication data set found for Barillas-Shanken (2018)'s 6-factor model
# We assemble it from several sources (as they apparently do)
## Fama-French's MKT.RF and SMB factors
bs.mktrf <- FF5.monthly$MKT.RF
bs.smb <- FF5.monthly$SMB
## Asness-Frazzini's HML.DEV and UMD (MOM) factors
source(file.path('inst', 'parsers', 'Value--Devil-in-HMLs-Details.R'))
bs.hml <- xts::xts(HML_Devil.HML.DEV$USA, HML_Devil.HML.DEV$DATE)
bs.mom <- xts::xts(HML_Devil.UMD$USA, HML_Devil.UMD$DATE)
rm(
  HML_Devil.HML_FF, HML_Devil.ME_1, HML_Devil.RF, 
  HML_Devil.MKT, HML_Devil.SMB, HML_Devil.HML.DEV, HML_Devil.UMD
)
## Hou-Xue-Zhang's IA and ROE factors
bs.ia <- Q4.monthly$IA
bs.roe <- Q4.monthly$ROE

# Our Barillas-Shanken data set
BS6.monthly <- merge(bs.mktrf, bs.smb, bs.hml, bs.mom, bs.ia, bs.roe)
BS6.monthly <- BS6.monthly[complete.cases(BS6.monthly), ]
colnames(BS6.monthly) <- c('MKT.RF', 'SMB', 'HML.DEV', 'MOM', 'IA', 'ROE')
rm(bs.mktrf, bs.smb, bs.hml, bs.mom, bs.ia, bs.roe)
```

It should be noted that there exist small nuances between the risk-free interest rates (one month T-Bill, `RF`) downloadable from French's website and those one can obtain from [global-q.org](http://global-q.org/index.html). These differences do not seem to be due to import conversions or numerical rounding, they may come directly from the source. 
In our context, they propagate to the market proxy factor (`MKT.RF`) and thus if anything we are forced to keep the two separate although they shall be conceptually the same depending on the assets universe considered.

# Q-factor models
Their *q-factor model* is inspired by the investment-based approach to the asset pricing theory.

## q-factor model 
In this section we first of all summarize research and the model introduced by \textcite{hou-al-2015}, then aim at reproducing it. It consists of four factors: the market factor ($MKT$), a size factor ($ME$), an investment factor ($I/A$), and a profitability factor ($ROE$). 
The factor regressions specification used to assess the q-factor model performance is 
$$
r_{i,t} - r_{f, t} = {\alpha}_{i,q} + {\beta}_{MKT,i}{MKT}_{t} + {\beta}_{ME,i}r_{ME,t} + {\beta}_{I/A,i}r_{I/A,t} + {\beta}_{ROE,i}r_{ROE,t} + {\epsilon}_{i,t}
$$
where, as usual, $r_{i,t} - r_{f, t}$ stands for the returns on excess of the risk-free rate.
Authors find their model to largely capture the cross section of average stock returns.

## $q^5$ factor model
In this section we first of all summarize research and the model introduced by \textcite{hou-al-2020}, then aim at reproducing it. 
This model is an extension of the previous four-factor q model in that it additionally includes the *expected growth* factor, $EG$.
The factor regressions specification used to assess the $q^5$-factor model performance is then 
$$
r_{i,t} - r_{f, t} = {\alpha}_{i,q} + {\beta}_{MKT,i}{MKT}_{t} + {\beta}_{ME,i}r_{ME,t} + {\beta}_{I/A,i}r_{I/A,t} + {\beta}_{ROE,i}r_{ROE,t} + {\beta}_{EG,i}r_{EG,t} + {\epsilon}_{i,t}
$$
# Benchmark models

## Fama-French five-factor model
The *Fama-French five-factor model* was introduced by \textcite{fama-french-2015}. In this model Fama and French include the RMW (Robust Minus Weak) and CMA (Conservative Minus Aggressive) to their standard \textcite{fama-french-1993} three-factor model. Factors are constructed using the six value-weight portfolios formed on size, the six value-weight portfolios formed on size and operating profitability, and the six value-weight portfolios formed on size and investment.

## Fama-French six-factor model
The *Fama-French six-factor model* is by \textcite{fama-french-2018}. It builds on the just discussed *Fama-French five-factor model* and adds the MOM momentum factor, which is also commonly referred to as Up Minus Down (UMD).
At this time we are unable to study the "alternative Fama-French six-factor model", which has the RMWc factor in place of the RMW factor, because authors apparently dismissed its publication. Readers willing to study this model would thus need to construct it themselves.

## Stambaugh-Yuan four-factor model
We provide a brief introduction to \textcite{stambaugh-yuan-2017} factors construction, focusing on their original contributed ones. In later sections the model is studied in comparison with the q-factor and q5 models, on a monthly basis. We share the factors monthly data set authors kindly made available on their website as an `.RData` object, further details can be found at `?SY4.monthly`.

Stambaugh-Yuan analyze a set of 11 anomalies. They categorize these anomalies in two clusters:
* _First cluster_: net stock issues, composite equity issues, accruals,
net operating assets, asset growth, and investment to assets.
* _Second cluster_: distress, O-score, momentum, gross profitability, and
return on assets.

Authors construct factors based on equally-weighted averages of stocks' anomaly rankings, in the perspective of having a less noisy mispricing measure for each stock across anomalies. In particular, stock's rankings are averaged with respect to the available anomaly measures within each of the two clusters. Thus, each month a stock has two composite mispricing measures, $P1$ and $P2$.
*Mispricing factors* are then constructed by applying a $2 \times 3$ sorting procedure, similarly to \textcite{fama-french-2015}:
* First, NYSE, AMEX, and NASDAQ stocks (excluding the ones with a price lower than 5\$) are sorted and split into two groups based on the NYSE median size breakpoint;
* Second, stock's are sorted by both $P1$ and $P2$ independently, and assigned to three groups ("low", "middle", and "high") with the 20th and 80th percentiles of the NYSE/AMEX/NASDAQ as breakpoints (rather than the commonly used 30th and 70th percentiles of the NYSE). The motivation authors provide for this methodological choice on breakpoints is that relative mispricing in the cross-section is considered to be "more a property of the extremes than of the middle".
* Finally, value-weighted returns of each of the four portfolios formed by the intersection of the two size categories with high and low categories of either $P1$ or $P2$ sorts are averaged and constitute their two mispricing factors, *MGMT* and *PERF*, respectively.

The \textcite{fama-french-2015}'s market proxy and size factors complete the four-factor model.

## Daniel-Hirshleifer-Sun three-factor model
We provide a brief introduction to \textcite{daniel-hirshleifer-sun-2020} factors construction, focusing on their original contributed ones. In later sections the model is studied in comparison with the q-factor and q5 models, on a monthly basis. We share the factors monthly data set authors kindly made available on their website as an `.RData` object, further details can be found at `?DHS3.monthly`.

Simplifying, \textcite{daniel-hirshleifer-sun-2020} factors construction is the following procedure:
* First, all NYSE, AMEX, and NASDAQ common stocks (CRSP 10 or 11 share codes, excluding financial firms and firms with negative book equity).
* Second, at June end firms are assigned to one of two size groups ("small" and "big"), depending on their ME being below or above the NYSE median size breakpoint.
* Finally, firms are also independently sorted into one of three financing groups ("low", "middle", "high") based on: either stocks' 1-year NSI and 5-year CSI financing measures rankings for the *FIN* factor, or the 4-day cumulative abnormal return around the most recent quarterly earnings announcement date (*CAR*) for the *PEAD* factor. Both sorts are with respect to NYSE 20th and 80th percentiles breakpoints.

The \textcite{fama-french-2015}'s market proxy factor completes the three-factor model.

## Barillas-Shanken six-factor model
We provide a brief introduction to \textcite{barillas-shanken-2018}. In later sections the model is studied in comparison with the q-factor and q5 models, on a monthly basis. To our knowledge there is not an out-of-the-box replication data set readily available for this model, but we rather assemble the model factors from other data sets available. At this time we are unable to share the complete data set as an `.RData` object to simply load into the environment.

In brief, the model six factors are: \textcite{fama-french-2015}'s market proxy excess returns (MKT.RF) and size (SMB) factors, \textcite{asness-frazzini-2013}'s "HML devil" (HML.DEV) and momentum (MOM, or equivalently UMD) factors, and finally \textcite{hou-al-2015}'s investment (IS) and profitability (ROE) factors.

# Models empirical comparisons
This section goal is twofold. First, closely following authors, we run spanning regressions in order to compare the q-factor models across the set of benchmark models considered and vice versa. Second, we compute a correlation matrix that includes the complete set of factors involved in all the models discussed.

Let us begin with the spanning regressions.

```{r SpanningRegressions utility function}
SpanningRegressions <- function(Y, X, ...) {
  Yn <- rep(Y, each=length(X)) 
  X <- rep(X, length(Y))
  # Models regressions
  fits <- mapply(
    function(y, x) {
      mod <- formula(
        paste(y, paste(x, collapse='+'), sep='~')
      )
      fit <- lm(mod, ...)
      return(fit)
    },
    as.list(Yn), X,
    SIMPLIFY=FALSE
  )
  # Newey-West t-statistics
  nwts <- lapply(fits, function(x) {
    lmtest::coeftest(x, vcov.=sandwich::NeweyWest(x, ...), ...)
  })
  # TODO: names overlap, workaround or longer names
  # Naming by regressand
  # names(fits) <- paste(Yn, 'reg', sep='.')
  # names(nwts) <- paste(Yn, 'nwts', sep='.')
  return(list(fits=fits, nwts=nwts))
}
```

```{r Table 1 Q factors vs Fama French}
q4.vars <- c('MKT.RF.1', 'ME', 'IA', 'ROE')
q5.vars <- c(q4.vars, 'EG')
ff5.vars <- c('MKT.RF', 'SMB', 'HML', 'RMW', 'CMA')
ff6.vars <- c(ff5.vars, 'MOM')
data.qff <- merge(FF6.monthly, Q5.monthly)
data.qff <- data.qff[complete.cases(data.qff), c(ff6.vars, q5.vars)]

# Panel A: Explaining q and q5 factors
q.on.ff <- SpanningRegressions(q5.vars, list(ff5.vars, ff6.vars), data=data.qff)
lapply(q.on.ff$fits, summary)

# Panel B: Explaining Fama–French factors
ff.on.q <- SpanningRegressions(ff6.vars, list(q4.vars, q5.vars), data=data.qff)
lapply(ff.on.q$fits, summary)
```

```{r Table 2 Q factors vs Stambaugh Yuan 4 Factors}
sy4.vars <- c('MKT.RF', 'SMB', 'MGMT', 'PERF')
data.qsy <- merge(SY4.monthly, Q5.monthly)
data.qsy <- data.qsy[complete.cases(data.qsy), c(q5.vars, sy4.vars)]

# Panel A: Explaining q and q5 factors
q.on.sy <- SpanningRegressions(q5.vars, list(sy4.vars), data=data.qsy)
lapply(q.on.sy$fits, summary)

# Panel B: Explaining the Stambaugh–Yuan factors
sy.on.q <- SpanningRegressions(sy4.vars, list(q4.vars, q5.vars), data=data.qsy)
lapply(sy.on.q$fits, summary)
```

```{r Table 3 Q factors vs Daniel Hirshleifer Sun 3 Factors}
dhs3.vars <- c('MKT.RF', 'PEAD', 'FIN')
data.qdhs <- merge(DHS3.monthly, Q5.monthly)
data.qdhs <- data.qdhs[complete.cases(data.qdhs), c(q5.vars, dhs3.vars)]

# Panel A: Explaining q and q5 factors
q.on.dhs <- SpanningRegressions(q5.vars, list(dhs3.vars), data=data.qdhs)
lapply(q.on.dhs$fits, summary)

# Panel B: Explaining the Daniel-Hirshleifer-Sun factors
dhs.on.q <- SpanningRegressions(dhs3.vars, list(q4.vars, q5.vars), data=data.qdhs)
lapply(dhs.on.q$fits, summary)
```

```{r Table 4 Q factors vs Barillas Shanken 6 Factors}
bs6.vars <- c('MKT.RF', 'SMB', 'HML.DEV', 'MOM', 'IA', 'ROE')
data.qbs <- merge(BS6.monthly, Q5.monthly)
data.qbs <- data.qbs[complete.cases(data.qbs), unique(c(q5.vars, bs6.vars))]

# Panel A: Regressing non-overlapping q5 factors on Barillas–Shanken factors
q5.on.bs <- SpanningRegressions(c('ME', 'EG'), list(bs6.vars), data=data.qbs)
lapply(q5.on.bs$fits, summary)

# Panel B: Regressing Asness–Frazzini HML factor on the q-factor and q5 models
hmldev.on.q <- SpanningRegressions(c('HML.DEV'), list(q4.vars, q5.vars), data=data.qbs)
lapply(hmldev.on.q$fits, summary)
```

Next, the correlation matrix, computed with factors series from July 1972 to December 2016.

```{r Table 5 Factors Correlation}
CorFactors <- function(data, ...) {
  # Correlation matrix of the full set of factors
  cor.estimates <- cor(data, ...)
  # Correlation coefficient test
  vars <- colnames(data)
  nvars <- length(vars)
  vars.pairs <- combn(vars, 2)
  cor.pvalues <- apply(vars.pairs, 2, function(j) {
    cor.test(data[, j[1]], data[, j[2]])$p.value
  })
  cor.pvalues <- matrix(cor.pvalues, nvars, nvars, byrow=TRUE)
  rownames(cor.pvalues) <- colnames(cor.pvalues) <- vars
  return(
    list(estimates=cor.estimates, pvalues=cor.pvalues)
  )
}

vars <- unique(c(q5.vars, ff6.vars, sy4.vars, dhs3.vars, bs6.vars))
data.all <- merge(FF6.monthly, Q5.monthly, SY4.monthly, DHS3.monthly, BS6.monthly)
data.all <- data.all[complete.cases(data.all), vars]
factors.correlation <- CorFactors(data.all)
```

